### 2015-09-29 manuscript.Rnw
starting to work on the examples

Generating phylogenies has been greatly facilitated by advances in sequencing
and computing technologies enabling the investigation of evolutionary
relationships among taxa. In parallel, new phylogenetic comparative methods have
been developed and make use of these phylogenies to explore fundamental
questions about the origin of biodiversity including the evolution of
morphological and ecological traits, the spatio-temporal variation in speciation
rates, or both \cite{Pennell2013}. The development of these methods has
contributed to the need for a greater availability of high-quality phylogenetic
information across life. However, phylogenetic information is scattered, often
only available as image files within publications, and the variety of file
formats used to store phylogenetic data makes it difficult for researchers to
access, synthesize, and integrate into their own research (\cite{Stoltzfus2012},
but see \cite{Cranston2014} for suggestions of best practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the Open Tree of Life contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore research questions in evolution, biodiversity, and conservation, the
resources OTL provides could be used in education and outreach (e.g.,
illustrating course material, develop outreach activities to explore
relationships among species).

@@

The Open Tree of Life project provides a web interface to explore, query and
download parts of the synthetic Open Tree
(\url{http://tree.opentreeoflife.org}). Additionally, the project provides an
Application Programming Interface (API) to the different services it offers
allowing users to work programmatically with the underlying data. Here, we
present \texttt{rotl} an R package that allows users to interact with OTL's API
to search for, and import phylogenetic trees and taxonomic information into
R. Having access to this data directly in R allows researchers to take advantage
of the multiple R packages that are available to visualize, analyze and
manipulate phylogenetic and comparative data (ape, phytools, geiger, ggtree,
phylobase, RNeXML, etc... cite task view).

@@

The OTL project provides four resources that serve data to users through the API
(the API endpoints):

\begin{enumerate}
\item The taxonomic information used as the backbone from the tree, the Open
  Tree Taxonomy;
\item The studies and their associated trees some of which are chosen by
  curators to assemble the synthetic tree;
\item A taxonomic name resolution service (TNRS) used to map taxon names to
  the Open Tree Taxonomy;
\item The synthetic tree itself (the Open Tree).
\end{enumerate}

@@

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used on the trees
returned by the API. However, resources and methods are being developed to add
branch lengths information to these topological trees (e.g., \cite{Ksepka2015})
or use topological trees to identify phylogenetically equivalent species to
increase overlap between chronograms and species trait data
\cite{Pennell2015}. Without branch lengths, these trees remain useful to
illustrate relationships among species, or to map a trait on a phylogeny.

\texttt{rotl} can also be used to download trees accompanying studies that have
been submitted through the curator interface, and identify the trees that are
being used to assemble the synthetic tree. These trees provide a useful resource
to reproduce or expand on an already published analysis.

@@

Use NCL/\texttt{rncl} to parse NEXUS/Newick files.

Taxonomic information, singletons, but NCL can parse it.

@@

\subsection{Getting tree from a given study}
\label{sec:get-tree-study}

* Which studies are available in OTL?
* How to look for them?

@@

>>>>

Generating phylogenies has been greatly facilitated by advances in sequencing
and computing technologies. In parallel, new phylogenetic comparative methods
have been developed and make use of these phylogenies to explore fundamental
questions about the origin of biodiversity including the evolution of
morphological and ecological traits, the spatio-temporal variation in speciation
rates, or both \cite{Pennell2013}. The development of these methods has
contributed to the need for a greater availability of high-quality phylogenetic
information across life. However, phylogenetic information is scattered, often
only available as image files within publications, and the variety of file
formats used to store phylogenetic data makes it difficult for researchers to
access, synthesize, and integrate this information into their own research
(\cite{Stoltzfus2012}, but see \cite{Cranston2014} for suggestions of best
practices).

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the Open Tree of Life contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore research questions in evolution, biodiversity, and conservation, the
resources OTL provides could be used for education and outreach (e.g.,
illustrating course material, develop outreach activities to explore
relationships among species).

@@

The Open Tree of Life project provides a web interface to explore, query and
download parts of the synthetic Open Tree
(\url{http://tree.opentreeoflife.org}). Additionally, the project provides an
Application Programming Interface (API) to the different services to work
programmatically with the underlying data. Here, we present \texttt{rotl} an R
package to interact with OTL's API using R syntax to search for, and import
phylogenetic trees and taxonomic information into R. Having access to this data
directly in R allows researchers to take advantage of the multiple packages that
are available to visualize, analyze and manipulate phylogenetic and comparative
data (ape, phytools, geiger, ggtree, phylobase, RNeXML, etc... cite task view
\url{https://cran.r-project.org/web/views/Phylogenetics.html}).

@@

The OTL project provides four resources that serve data to users through the
API:

\begin{enumerate}
\item The \emph{taxonomy} used as the backbone from the tree, the Open Tree
  Taxonomy;
\item The \emph{studies} and their associated trees some of which are chosen by
  curators to assemble the synthetic tree;
\item A \emph{taxonomic name resolution service} (TNRS) used to match taxon
  names to the Open Tree Taxonomy identifiers;
\item The \emph{synthetic tree} itself (the Open Tree).
\end{enumerate}

@@

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used directly on the
trees returned by the API. However, resources and methods are being developed to
add branch lengths information to these topological trees (e.g.,
\cite{Ksepka2015}) or use topological trees to identify phylogenetically
equivalent species to increase overlap between chronograms and species trait
data \cite{Pennell2015}. Without branch lengths, these trees remain useful to
illustrate relationships among species, or to map a trait on a phylogeny for
instance.

@@

Phylogenetic information retrieved from OTL is converted into
\texttt{ape::phylo} objects by \texttt{rotl} using the NEXUS Class Library
\citep[NCL,][]{Lewis2003} as implemented in the \texttt{rncl} package
\url{https://cran.r-project.org/package=rncl}. Using NCL provides robust and
efficient parsing of large trees that may contain singletons nodes labeled with
taxonomic information.

@@

To get the relationships among a set of taxa from the Open Tree, the taxa first
need to be matched against the Open Taxonomy using the TNRS. This step retrieves
the taxa identifiers that will be used to extract the relationships for the set
of requested taxa.

To illustrate how to obtain relationships from a set of taxa, we will rely on
OTL to draw a phylogenetic tree for a set of model organisms.

<<init, echo=FALSE>>=
library(rotl)
@

First, we use the function \texttt{tnrs\_match\_names} to match the taxon names to
their Open Taxonomy identifiers.

<<get_taxa, cache=TRUE>>=
taxa <- tnrs_match_names(names = c("Escherichia coli",
                                   "Chlamydomonas reinhardtii",
                                   "Drosophila melanogaster",
                                   "Arabidopsis thaliana",
                                   "Rattus norvegicus",
                                   "Mus musculus",
                                   "Cavia porcellus",
                                   "Xenopus laevis",
                                   "Saccharomyces cervisae",
                                   "Danio rerio"))
@

The function \texttt{tnrs\_match\_names} returns a data frame that lists the Open
Tree identifiers as well as other information to help users to ensure that the
taxa matched are the correct ones. Here, there is no ambiguity in the taxa
matched, however, as the Open Tree Taxonomy includes taxa from bacteria, plants,
and animals that are regulated by different nomenclatural codes (ICNP, ICN and
ICZN respectively), OTL and \texttt{rotl} provide tools to deal with potential
hemihomonyms. For instance, the argument \texttt{context\_name} can be used to
limit potential matches to a taxonomic group. The function
\texttt{tnrs\_contexts} provides a list of possible taxonomic contexts. When this
strategy cannot be used in case the tree encompasses multiple domains, the
functions \texttt{inspect} and \texttt{update} list alternative matches for a
taxon name, and replace it in the results, respectively. An example of this
approach is provided in the vignette ``How to use rotl?'' that accompanies the
package.

Now that the taxon names are matched to the Open Tree identifiers, we can pass
them to the function \texttt{tol\_induced\_subtree} to retrieve the
relationships among these taxa. In turn, the tree can be plotted directly as it
is returned as an \texttt{ape::phylo} object (Figure~\ref{fig:plot_taxa}).

<<plot_taxa, cache=TRUE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]])
plot(tree)
@

\subsection{Getting trees from studies}
\label{sec:get-tree-study}

\texttt{rotl} can also be used to retrieve trees accompanying studies that have
been submitted through the curator interface, and also identify the trees that
are being used to assemble the synthetic tree. These trees provide a useful
resource to reproduce or expand on an already published analysis.

Criteria that can be used to search for studies or their associated trees are
available through the output of the function
\texttt{studies\_properties}. Typically, users will want to search for studies
or trees based on a taxon but other criteria such as the title of the
publication. Here we demonstrate how to look for and retrieve a tree for study
focusing on cats, Felidae (Figure~\ref{fig:plot_cats}).

<<plot_cats, fig.cap="Tree from \\cite{} obtained from OTL">>=
cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
                                    value = "felidae")
cat_studies
cat_tree <- get_study_tree(study_id = cat_studies[["study_ids"]][1],
                           tree_id = list_trees(cat_studies)[[1]][1])
plot(cat_tree)
@

@@

\bibliographystyle{sysbio}

\bibliography{rotl-manuscript}
